Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-10 paper.qmd
moved maths definition stuff and context stuff down

## 2.1 Coloquial Definitions
The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## 2.2 Taxonomy Definitions
### 2.2.1 Taxonomies of Uncertainty
While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

@@

It is important to note that this taxonomy is the closest we can get to mathematical definition of uncertainty. While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* cannot be defined mathematically because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, whether mathematically defined or not. While the broader field of uncertainty visualisation does need to be more closely inspected, specifically through the scope of mathematical definitions, this issue cannot be solved with a more specific definition of uncertainty through mathematics. Instead or establishing a new definition of uncertainty that is strict enough to prevent the current array of misunderstandings, we should focus on inspecting how the current literature *deviates* from established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.

@@

@@

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017]. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

## 4.1 Similar work
There are a wealth of uncertainty communication and visualisation literature reviews that already exist.

@@

### A navigating a forest analogy
When communicating uncertainty information, it is important to ask if we should consider the bias to be variance. We often talk about the "bias variance trade off" but we rarely discuss bias as "cashed in" variance, which in some sense, it is. If you are struggling to understand how bias is just traded off variance, let us walk through an analogy.

Let us consider the data analysis pipeline a dense forest that you need to cross to get to your house. To prevent yourself from getting *too* lost you decide to bring some poles and rope so you can mark where you are going. You know your rope is long enough to get you to your house with enough wiggle room that you can reach the 5 houses either side of your house. This problem is drawn in @fig-forest1. @fig-forest2 depicts the typical data analysis pipeline and resulting uncertainty using this forest analogy. In this method, we navigate the forest by putting one of our poles down every time we notice the trees have changed (i.e. we have move to a new section of the forest). As we move down the line and we put our poles down, we cannot be certain we are always walking in the direction of our house, but we can do small checks or validation to make sure we are vaguely heading in the right direction. When we get to the final stage, we have enough rope remaining to get to the three houses directly in front of us, and we state that band as our "uncertainty". @fig-forest3 depicts the bias variance trade off in the case where we make no decisions, record as much information as possible, and check every single model with every single variable and keep every alternative as a possible option. This method will result in no bias (although it is often impossible to check the entirety of these spaces so it cannot) but a huge amount of variance.  In an effort to avoid putting down any poles, we have managed to navigate the forest, but have essentially no idea which of the 10 houses are ours. Of course, with sufficient feedback we can whittle down this vast array of choices (which is what a neural network can do) but feedback from an increased sample size can only influence variance, it cannot influence bias that was previously variance. @fig-forest4 shows an alternative way to perform the data analysis process to consider variance at stages beyond model uncertainty (such as the scenario uncertainty that comes with different human choices in climate change models). In this scenario you navigate the forest to a certain point, and then clone yourself three times and each put down a new pole at different distances to each other. All four versions of yourself will then walk in an identical fashion from the pole to the end of the forest, and each have your own path with your own confidence interval. This is how the [climate scenario uncertainty](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg) is constructed.

::: {#fig-forest layout="[[59,-2, 39], [59,-2, 39]]" layout-valign="bottom"}

![Data analysis Goal](forestwalk/fig1.jpg){#fig-forest1}

![The typical modelling process](forestwalk/fig2.jpg){#fig-forest2}

![Making zero assumption at every stage](forestwalk/fig3.jpg){#fig-forest3}

![An example of "scenario uncertainty" with a confidence interval](forestwalk/fig4.jpg){#fig-forest4}

An analogy of walking through the forest to help understand how our conceptualization of uncertainty and bias can influence what we consider to be uncertainty and what we consider to be bias. 
:::

This analogy may seem silly but it is explaining a very real problem that exists in uncertainty communication. The bias and variance trade off is commonly discussed in modelling, but rarely mentioned when we discuss communicating uncertainty. The reality is that every point in our data analysis, is full of decisions. Every time we make a decision at some point in this pipeline, we are impacting this bias and variance trade off. We have swapped out the variance from a large range of choices for the bias that may arise from an incorrect decision. Validation methods (such as cross validation, assumption checks, etc) help to guide the choices we make in the data analysis pipeline, but they can never offer any kind of certainty or truth considering validation methods themselves are also only estimates. When we consider the "uncertainty" of a model we usually only consider the variance of the final stage, the prediction error for a specific model and a specific set of features, and ignore the variance (that has been converted into bias by our selections) that was introduced at earlier stages (e.g. such as in data collection, when cleaning the data, or when considering alternative models). This is *well known* in modelling, it is why big tech companies try to collect as much data as they can and then put it in a neural network and hope they have enough data and enough feedback to avoid the this bias at *all* stages of the analysis. Recording *everything* means we don't even have to put any poles down at the data collection stage. Sometimes these statements are couched in a single assumption, such as when a time series model's confidence interval is constructed "assuming previous trends will continue into the future". This trade off issue means the uncertainty we typically estimate, in the form of a confidence inteval or traces or whatever you choose, has confidence intervals that are too small for what you actually consider the goal of the analysis to be. Unfortunately factors that typically decrease the variance, such as sample size, can only have an effect at the levels where that information is used to inform the choices. So confidence intervals that are incredibly small but built on a biased sample will be misleading and will be too small to *actually* predict future data. It is not uncommon for true results or models produced by a different team to sit entirely outside of the confidence band of another. This might be obvious when it is read, but the conversation about whether or not we should consider what was previously variance, to be variance when we communicate our analysis, is something that is rarely discussed although it is important.

Therefore, we must ask ourselves, *why* are we visualsing uncertainty? Uncertainty is only thought of as the final stages because to give a point prediction for a random process feels silly. By hiding the selections we made at earlier stages, when we traded that variance in for bias, we functionally made *point predictions* but it was just invisible to us.

@@

- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources. You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
# 4.?? Importance (and unimportance) of Secondary and Meta information
- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014

>>>>

## 2.1 Coloquial Definitions
::: aside
I am not sure how to collect these together and show that they are highly varied. This might be a situation for the tidytext clustering or something. There is an additional problem is that the definition for uncertainty is *completely* unrelated to the uncertainty that is quantified in the graphics. I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.
:::

The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## 2.2 Taxonomy Definitions
### 2.2.1 Taxonomies of Uncertainty
::: aside
I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon. That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with. The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work *that paper itself* identifies as transient.
:::

While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

@@

It is important to note that this taxonomy is thorough but it is not a mathematical definition of uncertainty. While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* has not been defined mathematically. This is largely because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, however this does not mean that a mathematical definition that encompases missing values, alternative models, and prediction intervals *cannot* be found. However, that task is far beyond the scope of this work. In this work, we will investigate the array of misunderstandings and focus on inspecting how the current literature *deviates* from already established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.

@@

::: aside
I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)
:::

@@

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

## 4.1 Similar work

There are a wealth of uncertainty communication and visualisation literature reviews that already exist.

@@

@@

- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources. You can project the uncertainty down into one dimension if the source information is in of itself unimportant.

# 4.?? Contextual Inormation
I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty. I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point. I actually think this quote from the grammar of graphics is exactly what I was trying to say. *"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect. Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991). But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*. Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain. The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data. *This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*. This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. The confusion that is created in the literature was going to be the topic I had later. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction. Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.

# 4.?? Importance (and unimportance) of Secondary and Meta information
- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014
